Keep your Laziness in Check

### 2018-03-06 paper/main.tex
[paper] more on section 3

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict (nf n) (if n > length xs then E d else endThunk d)
  where
    endThunk :: PosDemand [a] -> Demand [a]
    endThunk = \case
      Nil'        -> T
      dx :% T     -> E (dx :% T)
      dx :% E dxs -> E (dx :% endThunk dxs)
\end{inlinecode}
%
\hzh{TODO: describe what this spec means in English}

@@

\section{Strictness Specifications, by Example}
\label{sec:quickchick}

@@

There is a general pattern between a type \lk{a} and
its \textit{demand type}, for which we will write \lk{Demand a} from
now on. In both the example on lists and the example on maps,
a \lk{Thunk} type constructor is placed around all of the constructor
arguments to the original type. The \lk{Thunk} type has the following
definition:
\begin{inlinecode}
data Thunk a = E !a | T
\end{inlinecode}
Its two constructors \lk{E} and \lk{T} represents an evaluated
fragment of the original data, and unexplored regions of the original
data, respectively.

This design is motivated by the runtime representation of Haskell
values. Boxed values in Haskell are indirectly represented by a
pointer, and the pointer can lead to either a constructor with
pointers to deeper parts of the value, or a piece of unevaluated code,
which will produce a fragment of the value when evaluated. This piece
of code is also know as a thunk.

The \lk{Thunk} type reifies \hzh{there is a slight caveat here though,
we don't really capture sharing with this representation} the runtime
representation into a Haskell value. Programmers can then inspect
these demand values within Haskell. Most importantly, the demand types
form the vocabulary of specifications on laziness.

@@

StrictCheck specifications allows programmers to convey these
information accurately. A specification takes a demand on the output,
all of the inputs to the specified function, and computes a demand on
all the inputs. For example, the specification of the \lk{take}
function will have type \lk{Demand [a] -> Int -> [a] -> (Demand Int,
Demand [a])}.

\hzh{I want to say something about why we chose this design, but I'm not
actually sure why...}

The translation between a data type to its demand type and between a
function type to its specification type are mechanic, and although we
have presented them in a way that involves creating new demand type
definitions, our implementation handles all of this automatically
through generic programming. We delay this discussion
to \Cref{sec:generic-programming}.

>>>>

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict n (if n > length xs then d else d ++ thunk)
\end{inlinecode}
%
\hzh{TODO: describe what this spec means in English}

@@

\section{Strictness Specifications, by Example}
\label{sec:examples}

@@

There is a general pattern between a type \lk{a} and
its \textit{demand type}, for which we will write \lk{Demand a} from
now on. In both the example on lists and the example on maps,
a \lk{Thunk} type constructor is placed around all of the constructor
arguments to the original type. The \lk{Thunk} type has the following
definition:
\begin{inlinecode}
data Thunk a = E !a | T
\end{inlinecode}
Its two constructors \lk{E} and \lk{T} represents an evaluated
fragment of the original data, and unexplored regions of the original
data, respectively.
%
This design is motivated by the runtime representation of Haskell
values. Boxed values in Haskell are indirectly represented by a
pointer, and the pointer can lead to either a constructor bundled with
pointers to deeper parts of the value, or a piece of unevaluated code,
which will produce a fragment of the value upon execution. This piece
of code is also know as a thunk.

The \lk{Thunk} type reifies \hzh{there is a slight caveat here though,
we don't really capture sharing with this representation} the runtime
representation into a Haskell value. Programmers can then inspect
these demand values within Haskell. Most importantly, the demand types
form a critical part of specifications on laziness.

@@

StrictCheck specifications allows programmers to convey this
information accurately. In StrictCheck, specifications are represented
by the \lk{Spec} type constructor. The \lk{Spec} type constructor
takes two type arguments: the first is a type-level list holding the
types of arguments to specified function, and the second is the type
of the output of specified function. For example, consider the
function \lk{take :: Int -> [a] -> [a]}, one possible type of its
specification is \lk{Spec '[Int, [a]] [a]}.

In general, specification types are not unique for the same
function. Again consider \lk{take}, we can certainly choose the
specification type \lk{Spec '[Int] ([a] -> [a])} as well. However, the
meaning of this specification is very different from the other
choice. A specification of this type predicts how the \lk{Int}
argument is used given a demand on the partially-applied \lk{take}
function, and this specification does not produce information on how
the lists are used at all.

StrictCheck specifications apply to higher-order functions, and are
compositional. Take the example of \lk{map :: (a -> b) -> [a] -> [b]},
the type of its specification is \lk{Spec '[a] b -> Spec '[[a]] [b]},
a function from specifications to specifications. This allows
programmers to apply the specification of \lk{map} to any
specification of the function argument to \lk{map}, and conveniently
synthesize a specification based on that of \lk{map}.

The \lk{Spec} type also has a single constructor with the same
name. This constructor takes a higher-order function in
continuation-passing-style ~\cn, where the first argument to the
function is a framework-supplied function that takes the predicted
demand on the inputs. We will always name this argument \lk{predict}
in this paper. The function is CPS'd to hide some internal plumbing
that presents a variadic interface to the user, so that the number of
arguments to \lk{predict} always matches the length of the type-level
list in \lk{Spec}.

The translation between a data type to its demand type and between a
function type to its specification type are mechanic, and although we
have presented them in a way that involves creating new demand type
definitions in \Cref{sec:examples}, our implementation handles this
transition automatically through generic programming. We delay
discussion of this implementation detail
to \Cref{sec:generic-programming}.
